Unshuffling Permutations

### 2016-02-04 new-article.tex
LATIN 2016

\usepackage[utf8x]{inputenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{llncsdoc}
\usepackage[english]{babel}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{todonotes}
\usepackage{subfigure}
\usepackage{enumitem}
\usepackage{shuffle}
\usepackage{multirow}
\usepackage{esvect}

@@

\begin{abstract}
    A permutation is said to be a square if it can be obtained by
    shuffling two order-isomorphic patterns. The definition is intended
    to be the natural counterpart to the ordinary shuffle of words and
    languages. In this paper, we tackle the problem of recognizing square
    permutations from both the point of view of algebra and algorithmic.
    On the one hand, we present some algebraic and combinatorial
    properties of the shuffle product of permutations. We follow an
    unusual line consisting in defining the shuffle of permutations by
    means of an unshuffling operator, known as a coproduct. This
    strategy allows to obtain easy proofs for algebraic and combinatorial
    properties of our shuffle product. We besides exhibit a bijection
    between square $(213,231)$-avoiding permutations and square binary
    words. On the other hand, we prove that recognizing square
    permutations is \NPC. This property is proved by establishing a
    reduction from the pattern involvement problem and by using a
    containment-free perfect matching criterion to decide if a
    permutation is a square.
\end{abstract}

@@

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product} $\shuffle$ is a very famous operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word. In addition to purely combinatorial questions,
the shuffle product of words leads to the following computational ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there is a word $v$ such that $u$ is
    in $v \shuffle v$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
\begin{equation}
    O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
\begin{equation}
    O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
\end{equation}
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now finally focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. Hence, Problem~\ref{item:problem_4}
is \NPC.
\smallskip

This paper is intended to study a natural generalization of $\shuffle$,
denoted by $\SHUFFLE$, as a shuffle of permutations. Roughly speaking,
given three permutations $\pi$, $\sigma_1$, and $\sigma_2$, $\pi$ is
said to be a {\em shuffle} of $\sigma_1$ and $\sigma_2$ if $\pi$ (viewed
as a word) is a shuffle of two words that are order-isomorphic to
$\sigma_1$ and $\sigma_2$. Our intention in this paper is to study this
shuffle product of permutations $\SHUFFLE$ both from a combinatorial and
from a computational point of view by focusing on {\em square}
permutations, that are permutations $\pi$ being in the shuffle of a
permutation $\sigma$ with itself. The shuffle product defined and
considered in this paper is different from other existing and well-known
shuffle products on permutations. For instance, in~\cite{DHT:IJAC:2002},
the authors define the {\em shifted shuffle product}. For this product,
$\pi$ is a shuffle of $\sigma_1$ and $\sigma_2$ if $\pi$ is in the
shuffle, as words, of $\sigma_1$ and the word obtained by incrementing
all the letters of $\sigma_2$ by the size of $\sigma_1$. It is a simple
exercise to prove that, given three permutations $\pi$, $\sigma_1$, and
$\sigma_2$, deciding if $\pi$ is a in the shifted shuffle of $\sigma_1$
and $\sigma_2$ is in~\Pclass.
\smallskip

This paper is organized as follows. In
Section~\ref{section:Shuffle product on permutations} we provide a
precise definition of $\SHUFFLE$. This definition passes through the
preliminary definition of an operator~$\Delta$, allowing to
{\em unshuffle} permutations. This operator is in fact a coproduct,
endowing the linear span of all permutations with a coalgebra structure
(see~\cite{Joni:Rota:1979} or~\cite{Grinberg:Reiner:2014} for the
definition of these algebraic structures). By duality, the unshuffling
operator $\Delta$ leads to the definition of our shuffle operation on
permutations. This approach has many advantages. First, some
combinatorial properties of $\SHUFFLE$ depend on properties of $\Delta$
and are more easy to prove on the coproduct side. Second, this way of
doing allows to obtain a clear description of the multiplicities of the
elements appearing in the shuffle of two permutations, which are worthy
of interest from a combinatorial point of view.
Section~\ref{section:Binary square words and permutations} is devoted to
show that the problems related to the shuffle of words has links with
the shuffle of permutations. In particular, we show that binary words
that are square are in one-to-one correspondence with square
permutations avoiding some patterns
(Proposition~\ref{prop:bijection_binary_to_permutations_squares}). Next,
Section~\ref{section:Algebraic issues} presents some algebraic and
combinatorial properties of~$\SHUFFLE$. We show that $\SHUFFLE$ is
associative and commutative
(Proposition~\ref{prop:shuffle_associative_commutative}), and that if a
permutation is a square, its mirror, complement, and inverse are also
squares (Proposition~\ref{prop:square_stability}). Finally,
Section~\ref{section:Algorithmic issues} presents the most import result
of this paper: the fact that deciding if a permutation is a square is
\NPC (Proposition~\ref{proposition:hardness}). This result is obtained
by exhibiting a reduction from the pattern involvement
problem~\cite{Bose:Buss:Lubiw:1998} which is \NPC.

@@

We follow the usual terminology on words~\cite{ChoffrutKarhumaki1997}.
Let us recall here the most important ones. Let $u$ be a word. The
length of $u$ is denoted by $|u|$. The {\em empty word}, the only word
of null length, is denoted by $\epsilon$. We denote by $\widetilde{u}$
the {\em mirror image} of $u$ that is the word
$u_{|u|} u_{|u| - 1} \dots u_1$. If $P$ is a subset of $[|u|]$, $u_{|P}$
is the subword of $u$ consisting in the letters of $u$ at the positions
specified by the elements of $P$. If $u$ is a word of integers and $k$
is an integer, we denote by $u[k]$ the word obtained by incrementing by
$k$ all letters of $u$. The {\em shuffle} of two words $u$ and $v$ is
the set recursively defined by
$u \shuffle \epsilon = u = \epsilon \shuffle u$ and
$ua \shuffle vb = (u \shuffle vb)a \cup (ua \shuffle v)b$, were $a$ and
$b$ are letters. A word $u$ is a {\em square} if there exists a word $v$
such that $u$ belongs to $v \shuffle v$.

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$ are
specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations. The only difference lies on the
fact that we shall denote by $\pi(i)$ (instead of $\pi_i$) the $i$-th
letter of any permutation $\pi$. For any nonnegative integer $n$, we
write $\nearrow_{n}$ (resp. $\searrow_{n}$) for the permutation
$1 2 \dots n$ (resp. $n (n-1) \dots 1$). If $\pi$ is a permutation of
$S_n$, we denote by $\bar \pi$ the {\em complementary} of $\pi$ that is
the permutation satisfying $\bar \pi(i) = n - \pi(i) + 1$ for all
$i \in [n]$. The {\em inverse} of $\pi$ is denoted by $\pi^{-1}$.

If $u$ is a word of integers without multiple occurrences of a same
letter, $\STD(u)$ is the {\em standarized} of $u$, that is the unique
permutation of the same size as $u$ such that for all $i, j \in [|u|]$,
$u_i < u_j$ if and only if $\STD(u)(i) < \STD(u)(j)$. In particular, the
image of the map $\STD$ is the set $S$ of all permutations. Two words $u$
and $v$ having the same standarized are {\em order-isomorphic}. If
$\sigma$ is a permutation, there is an {\em occurrence} of (the
{\em pattern}) $\sigma$ in $\pi$ if there is a set $P$ of indexes of
letters of $\pi$ such that $\sigma$ and $\pi_{|P}$ are order-isomorphic.
When $\pi$ does not admit any occurrence of $\sigma$, $\pi$ {\em avoids}
$\sigma$.

Let us now provide some definitions about graphs and matching, used in
the sequel. For a graph $G$, we denote $\VS{G}$ as the set of vertices
and $\ES{G}$ as the set of edges of $G$. Let $u = u_1 u_2 \dots u_n$ be
a word. The \emph{clique graph associated to $u$}, denoted by $K_u$, is
the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The
structure of this underlying graph is linear, \emph{i.e.}, the set of
its vertices is equipped with a natural total order $<$ defined by
$u_i < u_j$ if and only if $i < j$. For a linear graph, we write
$(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
the context that $i < j$. Recall that two edges of a graph are
\emph{independent} if they do not share a common vertex, and that a
\emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
edges. A matching is \emph{perfect} if it covers all the vertices of the
graph. In case the set of vertices is equipped with a total order, a
matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
$\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$ or
$u_k < u_i < u_j < u_\ell$.

@@

Let use denote by $\QQ[S]$ the linear span of all permutations. We
define a linear coproduct $\Delta$ on $\QQ[S]$ in the following way. For
any permutation $\pi$, we set
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \sqcup P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2. 1 \otimes 12 +
    1 \otimes 21 + 2. 12 \otimes 1 + 21 \otimes 1 + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4. 1 \otimes 123 + 6. 12 \otimes 12 +
    4. 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3. {\bf 1 \otimes 132} + 1 \otimes 321 +
    3. 12 \otimes 21 \\ & + 3. 21 \otimes 12 + 3. 132 \otimes 1 +
    321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
Observe that the coefficient of the tensor $1 \otimes 132$ is $3$
in~\eqref{equ:example_unshuffling_coproduct} because there are exactly
three ways to extract from the permutation $1432$ two disjoint subwords
respectively order-isomorphic to the permutations $1$ and $132$.
\medskip

As announced, let us now use $\Delta$ to define a shuffle product on
permutations. As any coproduct, $\Delta$ leads to the definition of a
product obtained by duality in the following way.
From~\eqref{equ:unshuffling_coproduct}, for any permutation $\pi$, we
have
\begin{equation} \label{equ:schematic_coproduct}
    \Delta(\pi) =
    \sum_{\sigma, \nu \in S} \lambda_{\sigma, \nu}^\pi \;
    \sigma \otimes \nu,
\end{equation}
where the $\lambda_{\sigma, \nu}^\pi$ are nonnegative integers. Now,
by definition of duality, the dual product of $\Delta$, denoted by
$\SHUFFLE$, is a linear binary product on $\QQ[S]$. It satisfies, for
any permutations $\sigma$ and $\nu$,
\begin{equation}
    \sigma \SHUFFLE \nu =
    \sum_{\pi \in S}
    \lambda_{\sigma, \nu}^\pi \; \pi,
\end{equation}
where the coefficients $\lambda_{\sigma, \nu}^\pi$ are the ones
of~\eqref{equ:schematic_coproduct}. We call $\SHUFFLE$ the
{\em shuffle product of permutations}. For instance,
\begin{equation}\begin{split} \label{equ:example_shuffle_product}
    12 \SHUFFLE 21 & =
    1243 + 1324 + 2.1342 + 2.1423 + 3. {\bf 1432} + 2134 + 2.2314 \\
    & + 3.2341 + 2413 + 2.2431 + 2.3124 + 3142 + 3.3214 + 2.3241 \\
    & + 3421 + 3.4123 + 2.4132 + 2.4213 + 4231 + 4312.
\end{split}\end{equation}
Observe that the coefficient $3$ of the permutation $1432$
in~\eqref{equ:example_shuffle_product} comes from the fact that the
coefficient of the tensor $12 \otimes 21$ is $3$
in~\eqref{equ:example_unshuffling_coproduct}.
\medskip

Intuitively, this product shuffles the values and the positions of the
letters of the permutations. As first and easy properties of $\SHUFFLE$,
one can observe that the empty permutation $\epsilon$ is a unit for
$\SHUFFLE$ and that this product is graded by the sizes of the
permutations ({\em i.e.}, the product of a permutation of size $n$ with
a permutation of size $m$ produces a sum of permutations of size $n + m$).
Notice that there are other shuffle products on
permutations~\cite{DHT:IJAC:2002}: the {\em shifted shuffle} and the
{\em convolution product}. Our shuffle product is different from these
last two.
\medskip

We say that a permutation $\pi$ {\em appears} in the shuffle
$\sigma \SHUFFLE \nu$ of two permutations $\sigma$ and $\nu$ if the
coefficient $\lambda_{\sigma, \nu}^\pi$ defined above is different from
zero. In a more combinatorial way, this is equivalent to say that there
are two sets $P_1$ and $P_2$ of disjoints indexes of letters of $\pi$
satisfying $P_1 \sqcup P_2 = [|\pi|]$ such that the subword $\pi_{|P_1}$
is order-isomorphic to $\sigma$ and the subword $\pi_{|P_2}$ is
order-isomorphic to $\nu$.
\medskip

>>>>

\usepackage[utf8x]{inputenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{llncsdoc}
\usepackage[english]{babel}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{todonotes}
\usepackage{subfigure}
\usepackage{enumitem}
\usepackage{shuffle}
\usepackage{multirow}
\usepackage{esvect}
\usepackage{todonotes}
\usepackage[normalem]{ulem}

@@

\begin{abstract}
    A permutation is said to be a square if it can be obtained by
    shuffling two order-isomorphic patterns. The definition is intended
    to be the natural counterpart to the ordinary shuffle of words and
    languages. In this paper, we tackle the problem of recognizing square
    permutations from both the point of view of algebra and algorithmic.
    On the one hand, we present some algebraic and combinatorial
    properties of the shuffle product of permutations. We follow an
    unusual line consisting in defining the shuffle of permutations by
    means of an unshuffling operator, known as a coproduct. This
    strategy allows to obtain easy proofs for algebraic and combinatorial
    properties of our shuffle product. We besides exhibit a bijection
    between square $(213,231)$-avoiding permutations and square binary
    words. On the other hand, we prove that recognizing square
    permutations is \NPC. \sout{This property is proved by establishing a
    reduction from the pattern involvement problem and by using a
    containment-free perfect matching criterion to decide if a
    permutation is a square.}
    \uline{This property is proved by establishing a
    reduction from the pattern involvement problem and by using a
    contrained oriented perfect matching criterion to decide if a
    permutation is a square.}
\end{abstract}

@@

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The \emph{shuffle product} $\shuffle$ is a very famous operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word. In addition to purely combinatorial questions,
the shuffle product of words leads to the following computational ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there is a word $v$ such that $u$ is
    in $v \shuffle v$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
\begin{equation}
    O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
\begin{equation}
    O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
\end{equation}
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now finally focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. Hence, Problem~\ref{item:problem_4}
is \NPC.
\smallskip

This paper is intended to study a natural generalization of $\shuffle$,
denoted by $\SHUFFLE$, as a shuffle of permutations. Roughly speaking,
given three permutations $\pi$, $\sigma_1$, and $\sigma_2$, $\pi$ is
said to be a \emph{shuffle} of $\sigma_1$ and $\sigma_2$ if $\pi$ (viewed
as a word) is a shuffle of two words that are order-isomorphic to
$\sigma_1$ and $\sigma_2$. Our intention in this paper is to study this
shuffle product of permutations $\SHUFFLE$ both from a combinatorial and
from a computational point of view by focusing on \emph{square}
permutations, that are permutations $\pi$ being in the shuffle of a
permutation $\sigma$ with itself. The shuffle product defined and
considered in this paper is different from other existing and well-known
shuffle products on permutations. For instance, in~\cite{DHT:IJAC:2002},
the authors define the \emph{shifted shuffle product}. For this product,
$\pi$ is a shuffle of $\sigma_1$ and $\sigma_2$ if $\pi$ is in the
shuffle, as words, of $\sigma_1$ and the word obtained by incrementing
all the letters of $\sigma_2$ by the size of $\sigma_1$. It is a simple
exercise to prove that, given three permutations $\pi$, $\sigma_1$, and
$\sigma_2$, deciding if $\pi$ is a in the shifted shuffle of $\sigma_1$
and $\sigma_2$ is in~\Pclass.
\smallskip

This paper is organized as follows. In
Section~\ref{section:Shuffle product on permutations} we provide a
precise definition of $\SHUFFLE$. This definition passes through the
preliminary definition of an operator~$\Delta$, allowing to
\emph{unshuffle} permutations. This operator is in fact a coproduct,
endowing the linear span of all permutations with a coalgebra structure
(see~\cite{Joni:Rota:1979} or~\cite{Grinberg:Reiner:2014} for the
definition of these algebraic structures). By duality, the unshuffling
operator $\Delta$ leads to the definition of our shuffle operation on
permutations. This approach has many advantages. First, some
combinatorial properties of $\SHUFFLE$ depend on properties of $\Delta$
and are more easy to prove on the coproduct side. Second, this way of
doing allows to obtain a clear description of the multiplicities of the
elements appearing in the shuffle of two permutations, which are worthy
of interest from a combinatorial point of view.
Section~\ref{section:Binary square words and permutations} is devoted to
show that the problems related to the shuffle of words has links with
the shuffle of permutations. In particular, we show that binary words
that are square are in one-to-one correspondence with square
permutations avoiding some patterns
(Proposition~\ref{prop:bijection_binary_to_permutations_squares}). Next,
Section~\ref{section:Algebraic issues} presents some algebraic and
combinatorial properties of~$\SHUFFLE$. We show that $\SHUFFLE$ is
associative and commutative
(Proposition~\ref{prop:shuffle_associative_commutative}), and that if a
permutation is a square, its mirror, complement, and inverse are also
squares (Proposition~\ref{prop:square_stability}). Finally,
Section~\ref{section:Algorithmic issues} presents the most import result
of this paper: the fact that deciding if a permutation is a square is
\NPC (Proposition~\ref{proposition:hardness}). This result is obtained
by exhibiting a reduction from the pattern involvement
problem~\cite{Bose:Buss:Lubiw:1998} which is \NPC.

@@

We follow the usual terminology on words~\cite{ChoffrutKarhumaki1997}.
Let us recall here the most important ones. Let $u$ be a word. The
length of $u$ is denoted by $|u|$. The \emph{empty word}, the only word
of null length, is denoted by $\epsilon$. We denote by $\widetilde{u}$
the \emph{mirror image} of $u$ that is the word
$u_{|u|} u_{|u| - 1} \dots u_1$. If $P$ is a subset of $[|u|]$, $u_{|P}$
is the subword of $u$ consisting in the letters of $u$ at the positions
specified by the elements of $P$. If $u$ is a word of integers and $k$
is an integer, we denote by $u[k]$ the word obtained by incrementing by
$k$ all letters of $u$. The \emph{shuffle} of two words $u$ and $v$ is
the set recursively defined by
$u \shuffle \epsilon = u = \epsilon \shuffle u$ and
$ua \shuffle vb = (u \shuffle vb)a \cup (ua \shuffle v)b$, were $a$ and
$b$ are letters. A word $u$ is a \emph{square} if there exists a word $v$
such that $u$ belongs to $v \shuffle v$.

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$ are
specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations. The only difference lies on the
fact that we shall denote by $\pi(i)$ (instead of $\pi_i$) the $i$-th
letter of any permutation $\pi$. For any nonnegative integer $n$, we
write $\nearrow_{n}$ (resp. $\searrow_{n}$) for the permutation
$1 2 \dots n$ (resp. $n (n-1) \dots 1$). If $\pi$ is a permutation of
$S_n$, we denote by $\bar \pi$ the \emph{complementary} of $\pi$ that is
the permutation satisfying $\bar \pi(i) = n - \pi(i) + 1$ for all
$i \in [n]$. The \emph{inverse} of $\pi$ is denoted by $\pi^{-1}$.

If $u$ is a word of integers without multiple occurrences of a same
letter, $\STD(u)$ is the \emph{standarized} of $u$, that is the unique
permutation of the same size as $u$ such that for all $i, j \in [|u|]$,
$u_i < u_j$ if and only if $\STD(u)(i) < \STD(u)(j)$. In particular, the
image of the map $\STD$ is the set $S$ of all permutations. Two words $u$
and $v$ having the same standarized are \emph{order-isomorphic}. If
$\sigma$ is a permutation, there is an \emph{occurrence} of (the
\emph{pattern}) $\sigma$ in $\pi$ if there is a set $P$ of indexes of
letters of $\pi$ such that $\sigma$ and $\pi_{|P}$ are order-isomorphic.
When $\pi$ does not admit any occurrence of $\sigma$, $\pi$ \emph{avoids}
$\sigma$.

Let us now provide some definitions about graphs and \uline{oriented}
matching, used in the sequel.
For a graph $G$, we denote $\VS{G}$ as the set of vertices
and $\ES{G}$ as the set of edges of $G$. Let $u = u_1 u_2 \dots u_n$ be
a word. The \emph{clique graph associated to $u$}, denoted by $K_u$, is
the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The
structure of this underlying graph is linear, \emph{i.e.}, the set of
its vertices is equipped with a natural total order $<$ defined by
$u_i < u_j$ if and only if $i < j$. For a linear graph, we write
$(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
the context that $i < j$. Recall that two edges of a graph are
\emph{independent} if they do not share a common vertex, and that a
\emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
edges. A matching is \emph{perfect} if it covers all the vertices of the
graph.
\sout{In case the set of vertices is equipped with a total order, a
matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
$\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$ or
$u_k < u_i < u_j < u_\ell$.}
An \emph{oriented matching}
$\vv{\mathcal{M}}$ of $\mathcal{M}$ is a directed matching obtained
from $\mathcal{M}$ by orienting its edges.
\textbf{D\'efinir l'inclusion de motifs. En fait, je pense même qu'il ne faudrait
parler que de graphes orientés. Qu'en penses-tu?}

@@

Let use denote by $\QQ[S]$ the linear span of all permutations. We
define a linear coproduct $\Delta$ on $\QQ[S]$ in the following way. For
any permutation $\pi$, we set
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \sqcup P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the \emph{unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2. 1 \otimes 12 +
    1 \otimes 21 + 2. 12 \otimes 1 + 21 \otimes 1 + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4. 1 \otimes 123 + 6. 12 \otimes 12 +
    4. 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3. {\bf 1 \otimes 132} + 1 \otimes 321 +
    3. 12 \otimes 21 \\ & + 3. 21 \otimes 12 + 3. 132 \otimes 1 +
    321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
Observe that the coefficient of the tensor $1 \otimes 132$ is $3$
in~\eqref{equ:example_unshuffling_coproduct} because there are exactly
three ways to extract from the permutation $1432$ two disjoint subwords
respectively order-isomorphic to the permutations $1$ and $132$.
\medskip

As announced, let us now use $\Delta$ to define a shuffle product on
permutations. As any coproduct, $\Delta$ leads to the definition of a
product obtained by duality in the following way.
From~\eqref{equ:unshuffling_coproduct}, for any permutation $\pi$, we
have
\begin{equation} \label{equ:schematic_coproduct}
    \Delta(\pi) =
    \sum_{\sigma, \nu \in S} \lambda_{\sigma, \nu}^\pi \;
    \sigma \otimes \nu,
\end{equation}
where the $\lambda_{\sigma, \nu}^\pi$ are nonnegative integers. Now,
by definition of duality, the dual product of $\Delta$, denoted by
$\SHUFFLE$, is a linear binary product on $\QQ[S]$. It satisfies, for
any permutations $\sigma$ and $\nu$,
\begin{equation}
    \sigma \SHUFFLE \nu =
    \sum_{\pi \in S}
    \lambda_{\sigma, \nu}^\pi \; \pi,
\end{equation}
where the coefficients $\lambda_{\sigma, \nu}^\pi$ are the ones
of~\eqref{equ:schematic_coproduct}. We call $\SHUFFLE$ the
\emph{shuffle product of permutations}. For instance,
\begin{equation}\begin{split} \label{equ:example_shuffle_product}
    12 \SHUFFLE 21 & =
    1243 + 1324 + 2.1342 + 2.1423 + 3. {\bf 1432} + 2134 + 2.2314 \\
    & + 3.2341 + 2413 + 2.2431 + 2.3124 + 3142 + 3.3214 + 2.3241 \\
    & + 3421 + 3.4123 + 2.4132 + 2.4213 + 4231 + 4312.
\end{split}\end{equation}
Observe that the coefficient $3$ of the permutation $1432$
in~\eqref{equ:example_shuffle_product} comes from the fact that the
coefficient of the tensor $12 \otimes 21$ is $3$
in~\eqref{equ:example_unshuffling_coproduct}.
\medskip

Intuitively, this product shuffles the values and the positions of the
letters of the permutations. As first and easy properties of $\SHUFFLE$,
one can observe that the empty permutation $\epsilon$ is a unit for
$\SHUFFLE$ and that this product is graded by the sizes of the
permutations (\emph{i.e.}, the product of a permutation of size $n$ with
a permutation of size $m$ produces a sum of permutations of size $n + m$).
Notice that there are other shuffle products on
permutations~\cite{DHT:IJAC:2002}: the \emph{shifted shuffle} and the
\emph{convolution product}. Our shuffle product is different from these
last two.
\medskip

We say that a permutation $\pi$ \emph{appears} in the shuffle
$\sigma \SHUFFLE \nu$ of two permutations $\sigma$ and $\nu$ if the
coefficient $\lambda_{\sigma, \nu}^\pi$ defined above is different from
zero. In a more combinatorial way, this is equivalent to say that there
are two sets $P_1$ and $P_2$ of disjoints indexes of letters of $\pi$
satisfying $P_1 \sqcup P_2 = [|\pi|]$ such that the subword $\pi_{|P_1}$
is order-isomorphic to $\sigma$ and the subword $\pi_{|P_2}$ is
order-isomorphic to $\nu$.
\medskip

Besides, a permutation $\pi$ is a \emph{square} if there is a
permutation $\sigma$ such that $\pi$ appears in $\sigma \SHUFFLE \sigma$.
In this case, we say that $\sigma$ is a \emph{square root} of $\pi$.
Equivalently, $\pi$ is a square with $\sigma$ as square root if and only
if in the expansion of $\Delta(\pi)$, there is a tensor
$\sigma \otimes \sigma$ with a nonzero coefficient. In a more
combinatorial way, this is equivalent to say that there are two sets
$P_1$ and $P_2$ of disjoints indexes of letters of $\pi$ satisfying
$P_1 \sqcup P_2 = [|\pi|]$ such that the subword $\pi_{|P_1}$ and
$\pi_{|P_2}$ are order-isomorphic. Computer experiments give us the
first numbers of square permutations with respects to their size, which
are, from size $0$ to $10$,
\begin{equation}
    1, 0, 2, 0, 20, 0, 504, 0, 21032, 0, 1293418.
\end{equation}
This sequence (and its subsequence obtained by removing the $0$s) is for
the time being not listed in~\cite{Slo}. The square permutations of
sizes $0$ to $4$ are
\smallskip
